Estimation below the detection limit

### 2018-04-24 TEX/draft2.tex
Added figures as is: will edit text later
Updated Section 1

In this paper, we consider the problem of estimating a set of signals $x_1,\ldots,x_K$ from their multiple occurrences in unknown  locations in a data sequence $y$\TODO{This is the most important sentence of the paper, we need to polish it up}. The data may also contain background information---independent of the signals---modeled as noise.
For one-dimensional signals, the data sequence can be thought of as a 
long time series and the $K$ signals as repeated short events.  
This model appears in many applications, including spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.
In the last part of this manuscript, we also propose to interpreted this estimation problem as a toy model for the problem of single particle reconstruction using cryo-electron microscopy (cryo--EM). A precise mathematical formulation of the model and the estimation problem is provided in Section~\ref{sec:model}.

Figure~\ref{fig:XC_example} shows a simple example of data sequences with  one signal ($K=1$) that appears twice.
The upper row presents the data without noise, in the left panel, and its cross-correlation with the signal itself in the right panel. The cross-correlation exhibits two peaks corresponding to the signal's locations in the data. 
%Even if we add a small amount of noise and considering more than one signal ($K>1$), estimating the signals remains a painless task. 
%In this scenario, standard detection and clustering algorithms can locate and cluster the copies of the signals that can be then averaged within each class.  
In  the middle row, the same data is shown, now contaminated with i.i.d.\ Gaussian noise with zero mean and standard deviation of $\sigma=0.7$. 
While it is harder to observe the signal occurrences in the left panel, the cross-correlation still produces fairly adequate estimates of the signal's locations. In practice one usually does not possess a precise template of the signal to be recovered, however, clever methods based on template matching, such as those used in structural biology~\cite{heimowitz2018apple} and radar~\cite{gogineni2017passive}, may work.
The bottom panels show the same data swamped in  
i.i.d.\ Gaussian noise with standard deviation  $\sigma=3$. In this  low signal--to--noise (\SNR) regime, 
detection of individual signal occurrences is impossible, 
even if the true signal is known. 
This phenomenon---explained in more detail in Section~\ref{sec:model}--raises the question: 
\medskip 
\\
\emph{Can we estimate the signals accurately without detecting their individual occurrences? }
 \\

In this work, we provide a positive answer to this question. For one signal ($K=1$), we prove that in the asymptotic regime in which the signal appears
infinitely many times in the data, and under a spacing condition, one can estimate the signal \emph{to any desired accuracy in any $\SNR$ level}. We show empirically that the same holds true for multiple signals if $K$ is not too large. \TODO{Do we want a sentence like: Based on  recent results on the multireference alignment problem, we conjecture that it remains true as long as $K \lesssim L/6$. ?}

Our framework is based on autocorrelation analysis.
In a nutshell, the method consists of two stages. First, we estimate a mixture (linear combination) of the low--order autocorrelation functions of the signals from the data. These quantities can be estimated, to any desired accuracy, if individual occurrences are separated by at least $L-1$ entries and  each signal appears ample number of times \TODO{?} in the data. There is no need to detect individual occurrences. 
Furthermore, we do not assume the knowledge of the number of signal occurrences $M$ and the noise level.
In the second stage, the signals are estimated from the mixed autocorrelations using a nonconvex least-squares (LS).
Section~\ref{sec:autocorrelation} elaborates on the technique and proves some of its properties, while Section~\ref{sec:numerics} provides numerical demonstrations.

Interestingly, expectation-maximization (EM)---a popular framework for similar estimation problems, such as Gaussian mixture models and multireference alignment---is intractable for this problem, even if the number of signal occurrences $M$ is known and $K=1$.
In particular, in each iteration, EM assigns probabilities to all possible combinations of positioning the current signal estimate in $M$ locations on $\mathbb{Z}_N$, resulting in $O(N^M)$ possibilities (even when excluding forbidden combinations due to the spacing constraint).

\begin{figure}
		\centering
		\includegraphics[scale=0.5]{XC_example}
		\caption{ The upper left panel shows a data sequence of length $N=80$ with one rectangular signal of length $L=11$ that appears twice. In the second and third rows of the left column, a Gaussian noise with mean zero and standard deviation $\sigma=0.5$ and $\sigma=3$ was added, respectively. The panels in the right column present the cross-correlation of the data with the true rectangular signal. Note to the different scales in different panels. }
    	\label{fig:XC_example}
\end{figure}

@@

\subsection{One-dimensional experiments}

@@

\subsection{Two-dimensional experiments}

??

>>>>

In this paper, we consider the problem of estimating a set of signals $x_1,\ldots,x_K$ from their multiple occurrences in unknown  locations in a data sequence $y$\TODO{This is the most important sentence of the paper, we need to polish it up}. The data may also contain background information---independent of the signals---which we model as noise.
A precise mathematical formulation of the model and the estimation problem is provided in Section~\ref{sec:model}.
For one-dimensional signals, the data sequence can be thought of as a 
long time series and the $K$ signals as repeated short events.  
This model appears in many applications, including spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.
In the last part of this paper, we also propose to interpreted this estimation problem as a toy model for the task of single particle reconstruction using cryo-electron microscopy (cryo--EM).

Figure~\ref{fig:XC_example} shows a simple example of data sequences with  one signal ($K=1$).
The upper row presents the data without noise, in the left panel, and its cross-correlation with the signal itself in the right panel. The cross-correlation exhibits two peaks, corresponding to the two signal's locations in the data. 
%Even if we add a small amount of noise and considering more than one signal ($K>1$), estimating the signals remains a painless task. 
%In this scenario, standard detection and clustering algorithms can locate and cluster the copies of the signals that can be then averaged within each class.  
In  the middle row, the same data is shown, now contaminated with i.i.d.\ Gaussian noise with zero mean and standard deviation of $\sigma=0.5$. 
While it is harder to identify the signal occurrences in the left panel, the cross-correlation still produces fairly adequate estimates. In practice one usually does not possess a precise template of the signal to be recovered, however, clever methods based on template matching, such as those used in structural biology~\cite{heimowitz2018apple} and radar~\cite{gogineni2017passive}, may work.
The bottom panels show the same data swamped in  
i.i.d.\ Gaussian noise with standard deviation  $\sigma=3$. In this  low signal--to--noise (\SNR) regime, 
detection of individual signal occurrences is impossible, 
even if the true signal is known. 
This phenomenon---explained in more detail in Section~\ref{sec:model}--raises the following fundamental question: 
%\medskip 
%\\
\begin{center}
\emph{Can we estimate the signals accurately without detecting their individual occurrences? }  %\\
\end{center}

In this work, we provide a positive answer to this question. For a single signal ($K=1$), we prove that in the asymptotic regime in which the signal appears
infinitely many times, and under a spacing condition, one can estimate the signal \emph{to any desired accuracy in any $\SNR$ level}. We show empirically that the same holds true for multiple signals if $K$ is not too large. \TODO{Do we want a sentence like: Based on  recent results on the multireference alignment problem, we conjecture that it remains true as long as $K \lesssim L/6$. ? I am not sure it is necessary.}

Our framework is based on autocorrelation analysis.
In a nutshell, the method consists of two stages. First, we estimate a mixture (i.e., linear combination) of the low--order autocorrelation functions of the signals from the data. These quantities can be estimated, to any desired accuracy, if individual occurrences are separated by at least $L-1$ entries and  each signal appears sufficiently many times in the data. There is no need to detect individual occurrences. 
Furthermore, we do not assume the knowledge of the number of signal occurrences $M$ or the noise level.
In the second stage, the signals are estimated from the mixed autocorrelations using a nonconvex least-squares (LS).
Section~\ref{sec:autocorrelation} elaborates on the technique and proves some of its properties, while Section~\ref{sec:numerics} shows numerical demonstrations and discusses computational aspects.

Interestingly, expectation-maximization (EM)---a popular algorithm for similar estimation problems, such as Gaussian mixture models and multireference alignment---is intractable for this problem. This is true even if  $K=1$ and the number of signal occurrences $M$ is known.
In particular, in each iteration, EM assigns a probability to any feasible combination of positioning the current signal estimate in $M$ locations on the grid $\{1,\ldots,N\}$.
In total, even when excluding forbidden combinations due to the spacing constraint, there are $O(N^M)$ such combinations. \TODO{cumbersome}

\begin{figure}
		\centering
		\includegraphics[scale=0.5]{XC_example}
		\caption{ The upper left panel shows a data sequence (measurement) of length $N=80$ with one rectangular signal of length $L=11$ that appears twice. The upper right panel shows the cross-correlation of the data with the true rectangular signal.  
		The second and third rows present the data sequence with additive Gaussian noise with mean zero and standard deviation $\sigma=0.5$ and $\sigma=3$, respectively, and its cross-correlation with the signal. Note to the different scales in different panels. Importantly, in the low $\SNR$ regime the cross-correlation does not provide information on the locations of the signal in the data.}
    	\label{fig:XC_example}
\end{figure}

@@

%\subsection{One-dimensional experiments}

\TODO{Revise to explain $K = 1$ experiment first, then explain $K = 3$.}

\begin{figure}[t]
	\centering
	\includegraphics[width=0.7\linewidth]{XP_1D_homogeneous/progressive_n10000000000_107521}
	\caption{\TODO{...}}
	\label{fig:1Dhomosignals}
\end{figure}

\begin{figure}[t]
	\centering
	\includegraphics[width=0.7\linewidth]{XP_1D_homogeneous/progressive_RMSE_n10000000000_107521}
	\caption{\TODO{...}}
	\label{fig:1DhomoRMSE}
\end{figure}

@@

%\subsection{Two-dimensional experiments} 
%??
